Classical Planning Algorithms ¡ on the Atari Video Games

### 2014-11-19 icaps15-paper/paper.tex
Miquel's iteration

\newcommand{\Asterix}{\propername{Asterix}}
\newcommand{\BeamRider}{\propername{Beam Rider}}
\newcommand{\Breakout}{\propername{Breakout}}
\newcommand{\Enduro}{\propername{Enduro}}
\newcommand{\Freeway}{\propername{Freeway}}
\newcommand{\Pong}{\propername{Pong}}
\newcommand{\Qbert}{\propername{Q*Bert}}
\newcommand{\Seaquest}{\propername{Seaquest}}
\newcommand{\SpaceInvaders}{\propername{Space Invaders}}
\newcommand{\Deepmind}{\propername{Deepmind}}
\newcommand{\CrazyClimber}{\propername{Crazy\_climber}}
\newcommand{\Kangaroo}{\propername{Kangaroo}}
\newcommand{\Pooyan}{\propername{Pooyan}}
\newcommand{\Berzerk}{\propername{Berzerk}}
\newcommand{\Bowling}{\propername{Bowling}}
\newcommand{\Montezuma}{\propername{Montezuma\_revenge}}
\newcommand{\StarGunner}{\propername{Star\_gunner}}
\newcommand{\Venture}{\propername{Venture}}

@@

% \author{Nir Lipovetzky \\
%       The University of Melbourne \\
%    Melbourne, Australia\\
%    {\normalsize\url{@unimelb.edu.au}}
%    \And
%    Miquel Ramirez \\
%    RMIT University \\
%    Melbourne, Australia\\
%   {\normalsize\url{@rmit.edu.au}}
%    \And
%    Hector Geffner \\
%    ICREA \&  Universitat Pompeu Fabra \\
%    Barcelona, SPAIN \\
%    {\normalsize\url{@upf.edu}}\thanks{firstname.lastname}
% }

@@

\begin{abstract}
\begin{quote}
The Atari  2600 games supported in  the Arcade Learning Environment (Bellemare et. al. 2013)
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however,
cannot be used  for selecting actions  for  two reasons: first,  no  compact PDDL-model of  the games is given,
and  more importantly,  the action effects and goals are  not known \emph{a priori}. Moreover, in these games there is usually
no set of goals to be achieved but rewards to be collected. These features do not preclude the use of   classical  algorithms like
breadth-first search or Dijkstra's algorithm,  but    these methods are not effective over  large state spaces. We thus turn to a
different class of classical planning algorithms  introduced recently  that perform a \emph{structured exploration} of the state space; namely,
like breadth-first search and Dijkstra's  algorithm they are ``blind'' and hence  do not require prior knowledge of state transitions,
costs (rewards) or goals, and yet, like  heuristic search algorithms, they have been shown to be   effective for solving   problems over huge
state spaces. The simplest such algorithm, called Iterated Width or \IW, consists of  a sequence of calls  $\IW(1)$, $\IW(2)$, \ldots,
$\IW(k)$ where  $\IW(i)$ is  a breadth-first search in which a state is  pruned when it is not the
first state in the search  to make true some subset of $i$ atoms. 
The  empirical results over 35 games suggest that the performance 
of \IW\ with the  $k$ parameter fixed to $1$, i.e., $\IW(1)$, is at the level of the state of the art represented by 
\UCT.  A simple best-first variation  of \IW\  that combines exploration and exploitation proves
to be very competitive as well. 
\end{quote}
\end{abstract}

@@

The Iterated Width  (\IW) algorithm has been introduced as a classical planning algorithm
that takes a planning problem as an input, and computes an action sequence that solves the problem
as the output \cite{nir:ecai2012}. The algorithm however
applies to a broader range of problems. We will characterize such
problems  by means of  a finite and discrete set of states (the state space)
that correspond to vectors of size $n$. Namely, the
states are  \emph{structured} or \emph{factored} , and we take each of the locations in the vector
to represent a variable $X_i$, and the value at that vector location to
represent the value $x_j$ of variable $X_i$ in the state.
In addition to the state space, a problem is defined by an initial state $s_0$,
a set of actions applicable in each state, a
transition function $f$ such that  $s' = f(a,s)$ is
the state that results from applying action $a$  to the state $s$,
and rewards $r(a,s)$ represented by real numbers that result from  applying action $a$ in state $s$.
The transition and reward functions do not need to be known a priori, yet in that case, the
state and  reward that results from the application of an action in a state needs to be \emph{observable}.
The task is to compute an action sequence $a_0, \ldots, a_m$ for a large horizon $m$
that generates a state sequence $s_0, \ldots, s_{m+1}$ that  maximizes the accumulated
reward $\sum_{i=0}^{m} r(a_i,s_i)$, or that provides a good approximation.

@@

\IW\ is a systematic and complete blind-search algorithm like breadth-first search (\BRFS) and iterative deepening (ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the
search in a different way. This structured exploration  has proved to be very
effective over classical planning benchmark domains  when goals are  single atoms.\footnote{
Any conjunctive goal can be mapped into a single dummy atomic goal by adding an action that
achieves the dummy goal and that has the original conjunctive goal as a precondition.
Yet, this changes the definition of the domain.}
Table~\ref{table1} from \cite{nir:ecai2012} shows
the percentage of instances that are  solved by  the first iteration of \IW, i.e.  $\IW(1)$, by the second iteration
$\IW(2)$, and by neither one.  These are instances that  have been obtained from the existing benchmarks
by splitting problems with $N$ atomic goals, into $N$ problems with one atomic goal each. As the table shows,
$37\%$ of the 37921 instances are solved by $\IW(1)$ while $51.3\%$ are  solved by $\IW(2)$. Since $\IW(k)$
runs in time that is exponential in $k$, this mean that almost $90\%$ of the 37,921 instances are  solved in time that
is either linear or quadratic in the number of problem variables, which in these encoding are all \emph{boolean}.
Furthermore, when the performance of \IW\   is compared with  breadth-first search and iterative deepening,
on the one hand,  and  with a  Greedy Best First Search guided by the additive heuristic $h_{add}$
\cite{bonet:aij-hsp} on the other (this algorithm is similar
to the one used by  the FF planner when hill climbing search fails \cite{hoffmann:ff}),
it turns out that ``blind'' \IW\ solves as many problems as the informed search, 34,627 vs. 34,849, far ahead
of the other blind algorithms \BRFS\ and \ID\  that solve 9,010 and 8,762 problems each. This is shown in the bottom
part of  Table~\ref{table1}. Moreover, \IW\ is  faster and results in shorter plans than a heuristic search algotrithm
\cite{nir:ecai2012}.

@@

\input{games-table}
We tested \IW(1) and \BFS\ over $35$ of the 55 games considered in
\cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
Entries marked as  ``--'' in the table are   still running at the time of submission. 
Each run can take up to 24 hours of CPU time and there are 5 runs per game and algorithm. 
Eventually all of  the  55 games will  be included in the evaluation.}
Table~\ref{performance} shows the performance of $\IW(1)$ and \BFS\
% , characterized by the average score obtained over all episodes, the standard deviation
% of the score obtained and time (in seconds) per lookahead,
in comparison  with breadth-first search (\BRFS) and \UCT.
Videos of selected games played by \IW(1), \BFS, and \UCT\ can be 
seen in Youtube.\footnote{\url{http://www.youtube.com/playlist?list=PLXpQcXUQ_CwenUazUivhXyYvjuS6KQOI0}.}
The discount factor used by  all the algorithms is $\gamma=0.995$.
The scores for \BRFS\ and \UCT\ are from BNVB.  
Our experimental setup  follows theirs except that a maximum budget of  $150,000$ simulated frames 
is applied to \textit{all the algorithms}. \UCT\ uses this budget by running
 $500$   rollouts of depth $300$. % \footnote{In \UCT,  $500$ base policy rollouts
% of depth $300$ is equivalent to a budget of $150,000$ simulated frames.}.
The bound on the number of simulated frames is like a bound on
lookahead time, as most of the time in the lookahead  is spent
in calls to the emulator for computing the next RAM state. 
This is why the average time per action is similar to all the algorithms
except \IW(1), that due to its agressive pruning does not always use
the full budget and takes less time per action on average.

Also,  as reported by  BNVB, all of the algorithms 
reuse the subtree from the previous lookahead that is rooted in the selected 
child, deleting its siblings and their descendants. More precisely, no calls 
to the emulator are done for transitions that are cached in that subtree, 
and such reused frames are not discounted from the budget that is thus a bound 
on the number of ``new'' frames per lookahead. In addition, in \IW(1), the states that 
are ``reused'' from the previous searches are ignored in the computation of the novelty of 
new states so that more states can escape pruning. Otherwise, \IW(1) often
uses a fraction of the  budget.  This is not needed in \BFS\ which does 
no pruning.  \IW(1) and \BFS\ are  limited to  search up to a depth of
$m=1500$ frames and up to $m=150,000$ frames per root branch. This is
to avoid the search from  going too deep or being too committed to a
single root action. 
% Similarly UCT defines a limit of $m=300$ frames for each rollout. 
% Note that UCT is more likely to have a performance
% loss if the depth is increased, as the number of rollouts would
% decrease in order to maintain the same budget of $150,000$ simulated
% frames.

Last, \IW(1) and \BFS\ select an action every 5 frames, while UCT  selects 
an action in every frame.  This means that in order to explore a branch 
300 frames deep, UCT gets to choose 300 actions, while \IW(1) and \BFS\ 
get to choose 60 actions, both however using the same 300 frames from the
budget. For this, we  followed the setup  of \BRFS\ in BNVB 
that also selects  actions every 5 frames. Since the lookahead budget 
is given by a maximum number of (new) frames, and the time is mostly taken by 
calls to the emulator, this may not be the best choice  for \IW(1)\ 
and \BFS\ that are therefore not exploiting all the options afforded
by the budget. We hope to fix this gap in the future.

\Omit{
The last difference inherited from BNVB between all algorithms
and UCT, is that UCT selects an action randomly in the rollout every
single frame, while the others select an action every 5 frames, and
the remain ones are $no-op$, as is done in the simulation
environment. Therefore, a node in all algorithms represent $5$ frames,
but in UCT represents $1$ frame. As a result, a UCT rollout of depth
$300$ frames involve the selection of 300 actions, decreasing the
budget of simulated frames by $300$, while the other algorithms need a
branch of $1500$ frames to select $300$ actions, decreasing the budget
by $1500$ frames instead.
}

Table~\ref{performance} shows that both $\IW(1)$ and \BFS\  outperform
\BRFS, which rarely collects any reward in many domains, as the
depth of the \BRFS\ search tree results in a lookahead of $0.3$
seconds ($20$ frames or $4$ nodes deep). On the other hand, both
$\IW(1)$ and \BFS\ reach states that are up to $350$--$1500$ frames
deep ($70$--$260$ nodes or $6$--$22$ seconds), even if $\IW(1)$
doesn't always use all the simulation frames allocated due to its
pruning. This behaviour can be observed in games such as
\Breakout, \CrazyClimber, \Kangaroo, and \Pooyan, where the average
CPU time for each lookahead is up to $10$ times faster than \BFS. 
Computation time for \UCT\ and \BRFS\ are  similar to \BFS, as the
most expensive part of the computation is the  generation of 
frames through the simulator,  and these three  algorithms 
always use  the full budget.

More interestingly,  \IW(1) outscores \UCT\  in 17 of the 35 games,
while \BFS\ outscores \UCT\ in 18.\footnote{These numbers can actually
improve further as 6  games for \IW(1) and 4 games for \BFS\ are still running.}
On the other, \UCT\ does better than \IW(1) and \BFS\ in 9 and 11 games respectively.
The relative performance between \IW(1)\ and \BFS\ is closer with \IW(1) being  the
best of the two in 12 games, and  \BFS\  in 11. In terms of the number of games 
where an algorithm is the overall winner, the situation is  even: 
\IW(1) is the overall best in 12 games, \BFS\ in 11 games, and \UCT\ in 12 games. 
Also, \BRFS\ is best in 1 game (Centipede), while the other three algorithms
are tied in another game (\Pong).

@@

$\IW(1)$ also outperforms the best learning algorithm in
\cite{deep-mind-atari} in the $7$ games considered, and
\BFS\ does so in $6$ of the $7$ games. Likewise, in \Freeway, \Berzerk, \Bowling, \Montezuma,
\StarGunner, and \Venture, both $\IW(1)$ and \BFS\ attain a better
score than the baseline semi-random algorithm \emph{Perturb} in
\cite{bellemare:jair2013}.
%% to say the 2 lines below must add context: .. eg
%%% ``which in *these games* outperforms both \UCT\ and the reinforcement learning methods.''
%% add it if that's the case ..
\emph{Perturb} is a simple algorithm
that selects  a fixed  action with probability 0.95, and a random
action with probability 0.05.  For \emph{Perturb}, BNVB do not report 
the average score  but the  best score.  \emph{Perturb} manages to do well in domains where rewards
are deep but can be  reached by  repeating the same action.
This is the case of \Freeway, where a chicken has to run across a ten lane
highway filled with traffic. Every time the chicken gets across, there is one
unit of reward. If the chicken is hit by a car, it has to start again. 
In \Freeway, 12 out of 18 possible actions have an effect:
6 actions behave as move-up (up-right, up-left, up-fire, up-right-fire, up-left-fire) , 
6 as  move-down (down-right, down-left, down-fire, down-right-fire, down-left-fire), 
and 6 actions do nothing.  \emph{Perturb}  does well in this domain 
when the selected fixed action is move-up.  As seen 
in Table~\ref{performance}, \UCT\ does not manage to take any chicken
across the highway. The reason that  \UCT\ performs so erratically 
and the rollouts do not manage to collect any reward, is that there  is a $1/3$ probability of  
choosing a move up action,  and   200 move up actions are required  in order to collect a reward. 
%% the previous number was not correct; no-op actions do not count, and thus needs
%%% 200 more move up actions than move down actions, although this is not accurate either
%%%% as move down actions that take chicken below the screen count as no-ops too ..
\IW\ does not have this limitation and is best in \Freeway.

@@

Another challenging game is \Montezuma. In this game rewards are very
sparse, deep, and most of the actions lead to losing a  life with
no immediate penalty or consequence.  All algorithms achieve 0 score, 
except for \BFS\ that achieves an average score of 80, and a score
of 400 in one of the runs. This means that even \BFS\  is not able to 
consistently find rewards in this game. This game and several other games 
like \Breakout, and \SpaceInvaders\ could be much simpler by adding
negative rewards for losing a life. We have indeed observed that
our planning algorithms don't care much about losing lifes
until there is just one life left, when they play much better.
This can be seen in the videos mentioned above, and points
to a simple form of learning that would be useful both
for planners and reinforcement learning approaches.

@@

We have shown experimentally that width-based algorithms like $\IW(1)$ and \BFS\
that originate in work in classical planning,  can be used to play  the Atari video
games where they  achieve  state-of-the-art performance. This level of play 
is the result of a structured exploration of the state space that manages to combine 
the scope of blind search algorithms with the efficiency of heuristic search methods. 
There are several  issues worth studying, we  mention two of them.
Regarding planning, the results show the potential of width-based algorithms
for a broader class of planning problems. As shown, indeed, these algorithms
do not require PDDL encodings and can deal with state successor functions
encoded procedurally (as in simulators) as long as the structure of the states 
is known. Regarding the Atari games, two problems of  all the  planning methods
in ALE are first,  that by using the RAM state of the Atari console they are ``cheating'', 
as they use information that is not available to the human player, and second, 
that they do not get better with experience. In the future, we would like to explore 
whether  variations of these algorithms  can be used to play  from the information in the  
screen state as opposed to the RAM state, and potential uses of the  new ideas
for learning to improve performance.

>>>>

\newcommand{\numgames}{$35$}

\newcommand{\Assault}{\propername{Assault}}
\newcommand{\Asterix}{\propername{Asterix}}
\newcommand{\Asteroids}{\propername{Asteroids}}
\newcommand{\Atlantis}{\propername{Atlantis}}
\newcommand{\BankHeist}{\propername{Bank Heist}}
\newcommand{\BattleZone}{\propername{Battle Zone}}
\newcommand{\BeamRider}{\propername{Beam Rider}}
\newcommand{\Berzerk}{\propername{Berzerk}}
\newcommand{\Bowling}{\propername{Bowling}}
\newcommand{\Boxing}{\propername{Boxing}}
\newcommand{\Breakout}{\propername{Breakout}}
\newcommand{\Centipede}{\propername{Centipede}}
\newcommand{\ChopperCommand}{\propername{Chopper Command}}
\newcommand{\CrazyClimber}{\propername{Crazy Climber}}
\newcommand{\Enduro}{\propername{Enduro}}
\newcommand{\Freeway}{\propername{Freeway}}
\newcommand{\Frostbite}{\propername{Frostbite}}
\newcommand{\Gravitar}{\propername{Gravitar}}
\newcommand{\Hero}{\propername{Hero}}
\newcommand{\JamesBond}{\propername{James Bond}}
\newcommand{\Kangaroo}{\propername{Kangaroo}}
\newcommand{\Krull}{\propername{Krull}}
\newcommand{\Montezuma}{\propername{Montezuma Revenge}}
\newcommand{\MsPacman}{\propername{Ms Pacman}}
\newcommand{\Pong}{\propername{Pong}}
\newcommand{\Pooyan}{\propername{Pooyan}}
\newcommand{\Qbert}{\propername{Q*Bert}}
\newcommand{\RiverRaid}{\propername{Riverraid}}
\newcommand{\Seaquest}{\propername{Seaquest}}
\newcommand{\SpaceInvaders}{\propername{Space Invaders}}
\newcommand{\StarGunner}{\propername{Star Gunner}}
\newcommand{\Tennis}{\propername{Tennis}}
\newcommand{\Venture}{\propername{Venture}}
\newcommand{\VideoPinball}{\propername{Video Pinball}}
\newcommand{\Zaxxon}{\propername{Zaxxon}}

\newcommand{\Deepmind}{\propername{Deepmind}}

@@

% \author{Nir Lipovetzky \\
%       The University of Melbourne \\
%    Melbourne, Australia\\
%    {\normalsize\url{@unimelb.edu.au}}
%    \And
%    Miquel Ramirez \\
%    Australian National University \\
%    Canberra, Australia\\
%   {\normalsize\url{@gmail.com}}
%    \And
%    Hector Geffner \\
%    ICREA \&  Universitat Pompeu Fabra \\
%    Barcelona, SPAIN \\
%    {\normalsize\url{@upf.edu}}\thanks{firstname.lastname}
% }

@@

\begin{abstract}
\begin{quote}
The Atari 2600 games supported in  the Arcade Learning Environment~\cite{bellemare:jair2013}
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however,
cannot be used  for selecting actions  for  two reasons: first,  no  compact PDDL-model of  the games is given,
and  more importantly,  the action effects and goals are  not known \emph{a priori}. Moreover, in these games there is usually
no set of goals to be achieved but rewards to be collected. These features do not preclude the use of   classical  algorithms like
breadth-first search or Dijkstra's algorithm,  but    these methods are not effective over  large state spaces. We thus turn to a
different class of classical planning algorithms  introduced recently  that perform a \emph{structured exploration} of the state space; namely,
like breadth-first search and Dijkstra's  algorithm they are ``blind'' and hence  do not require prior knowledge of state transitions,
costs (rewards) or goals, and yet, like  heuristic search algorithms, they have been shown to be   effective for solving   problems over huge
state spaces. The simplest such algorithm, called Iterated Width or \IW, consists of  a sequence of calls  $\IW(1)$, $\IW(2)$, \ldots,
$\IW(k)$ where  $\IW(i)$ is  a breadth-first search in which a state is  pruned when it is not the
first state in the search  to make true some subset of $i$ atoms. 
The  empirical results over 35 games suggest that the performance 
of \IW\ with the  $k$ parameter fixed to $1$, i.e., $\IW(1)$, is at the level of the state of the art represented by 
\UCT.  A simple best-first variation  of \IW\  that combines exploration and exploitation proves
to be very competitive as well. 
\end{quote}
\end{abstract}

@@

The Iterated Width  (\IW) algorithm has been introduced as a classical planning algorithm
that takes a planning problem as an input, and computes an action sequence that solves the problem
as the output \cite{nir:ecai2012}. The algorithm however
applies to a broader range of problems. We will characterize such
problems  by means of  a finite and discrete set of states (the state space)
that correspond to vectors of size $n$. Namely, the
states are  \emph{structured} or \emph{factored} , and we take each of the locations in the vector
to represent a variable $X_i$, and the value at that vector location to
represent the value $x_j$ of variable $X_i$ in the state.
In addition to the state space, a problem is defined by an initial state $s_0$,
a set of actions applicable in each state, a
transition function $f$ such that  $s' = f(a,s)$ is
the state that results from applying action $a$  to the state $s$,
and rewards $r(a,s)$ represented by real numbers that result from  applying action $a$ in state $s$.
The transition and reward functions do not need to be known \emph{a priori}, yet in that case, the
state and  reward that results from the application of an action in a state needs to be \emph{observable}.
The task is to compute an action sequence $a_0, \ldots, a_m$ for a large horizon $m$
that generates a state sequence $s_0, \ldots, s_{m+1}$ that  maximizes the accumulated
reward $\sum_{i=0}^{m} r(a_i,s_i)$, or that provides a good approximation.

@@

\IW\ is a systematic and complete blind-search algorithm like breadth-first search (\BRFS) and iterative deepening (\ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the
search in a different way. This structured exploration  has proved to be very
effective over classical planning benchmark domains  when goals are  single atoms.\footnote{
Any conjunctive goal can be mapped into a single dummy atomic goal by adding an action that
achieves the dummy goal and that has the original conjunctive goal as a precondition.
Yet, this changes the definition of the domain.}
Table~\ref{table1} from \cite{nir:ecai2012} shows
the percentage of instances that are  solved by  the first iteration of \IW, i.e.  $\IW(1)$, by the second iteration
$\IW(2)$, and by neither one.  These are instances that  have been obtained from the existing benchmarks
by splitting problems with $N$ atomic goals, into $N$ problems with one atomic goal each. As the table shows,
$37\%$ of the 37921 instances are solved by $\IW(1)$ while $51.3\%$ are  solved by $\IW(2)$. Since $\IW(k)$
runs in time that is exponential in $k$, this mean that almost $90\%$ of the 37,921 instances are  solved in time that
is either linear or quadratic in the number of problem variables, which in these encoding are all \emph{boolean}.
Furthermore, when the performance of \IW\   is compared with  breadth-first search and iterative deepening,
on the one hand,  and  with a  Greedy Best First Search guided by the additive heuristic $h_{add}$
\cite{bonet:aij-hsp} on the other (this algorithm is similar
to the one used by  the FF planner when hill climbing search fails \cite{hoffmann:ff}),
it turns out that ``blind'' \IW\ solves as many problems as the informed search, 34,627 vs. 34,849, far ahead
of the other blind algorithms \BRFS\ and \ID\  that solve 9,010 and 8,762 problems each. This is shown in the bottom
part of  Table~\ref{table1}. Moreover, \IW\ is  faster and results in shorter plans than a heuristic search algorithm
\cite{nir:ecai2012}.

@@

\input{games-table}
We tested \IW(1) and \BFS\ over \numgames\
 of the $55$ games considered in
\cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
Entries marked as  ``--'' in the table are   still running at the time of submission. 
Each run can take up to $24$ hours of CPU time and there are $10$ runs per game and algorithm. 
Eventually all of  the  $55$ games will  be included in the evaluation.}
Table~\ref{performance} shows the performance of $\IW(1)$ and \BFS\
% , characterized by the average score obtained over all episodes, the standard deviation
% of the score obtained and time (in seconds) per lookahead,
in comparison  with breadth-first search (\BRFS) and \UCT.
Videos of selected games played by $\IW(1)$, \BFS, and \UCT\ can be 
seen in Youtube.\footnote{\url{http://www.youtube.com/playlist?list=PLXpQcXUQ_CwenUazUivhXyYvjuS6KQOI0}.}
The discount factor used by  all the algorithms is $\gamma=0.995$.
The scores reported for \BRFS\ and \UCT\ are taken from BNVB.  
Our experimental setup  follows theirs except that a maximum budget of  $150,000$ simulated frames 
is applied to \textit{all the algorithms}. \UCT\ uses this budget by running
 $500$   rollouts of depth $300$. % \footnote{In \UCT,  $500$ base policy rollouts
% of depth $300$ is equivalent to a budget of $150,000$ simulated frames.}.
The bound on the number of simulated frames is like a bound on
lookahead time, as most of the time in the lookahead  is spent
in calls to the emulator for computing the next RAM state. 
This is why the average time per action is similar to all the algorithms
except $\IW(1)$, that due to its aggressive pruning does not always use
the full budget and takes less time per action on average.

Also,  as reported by  BNVB, all of the algorithms 
reuse the sub-tree from the previous lookahead that is rooted in the selected 
child, deleting its siblings and their descendants. More precisely, no calls 
to the emulator are done for transitions that are cached in that sub-tree, 
and such reused frames are not discounted from the budget that is thus a bound 
on the number of ``new'' frames per lookahead. In addition, in $\IW(1)$, the states that 
are ``reused'' from the previous searches are ignored in the computation of the novelty of 
new states so that more states can escape pruning. Otherwise, $\IW(1)$ often
uses a fraction of the  budget.  This is not needed in \BFS\ which does 
no pruning.  $\IW(1)$ and \BFS\ are  limited to  search up to a depth of
$m=1,500$ frames and up to $m=150,000$ frames per root branch. This is
to avoid the search from  going too deep or being too committed to a
single root action. 
% Similarly UCT defines a limit of $m=300$ frames for each rollout. 
% Note that UCT is more likely to have a performance
% loss if the depth is increased, as the number of rollouts would
% decrease in order to maintain the same budget of $150,000$ simulated
% frames.

Last, \IW(1) and \BFS\ select an action every $5$ frames, while \UCT\  selects 
an action in every frame.  This means that in order to explore a branch 
$300$ frames deep, UCT gets to choose $300$ actions, while \IW($1$) and \BFS\ 
get to choose $60$ actions, both however using the same $300$ frames from the
budget. For this, we  followed the setup  of \BRFS\ in BNVB 
that also selects actions every $5$ frames, matching the behavior
of the emulator that requests an action also every $5$ frames. Since the 
lookahead budget is given by a maximum number of (new) frames, and the time 
is mostly taken by calls to the emulator, this may not be the best choice  
for $\IW(1)$ and \BFS\ that may therefore not be exploiting all the 
options afforded by the budget. We hope to investigate the implications of
this gap in the future.

\Omit{
The last difference inherited from BNVB between all algorithms
and \UCT\, is that \UCT\ selects an action randomly in the rollout every
single frame, while the others select an action every 5 frames, and
the remain ones are $no-op$, as is done in the simulation
environment. Therefore, a node in all algorithms represent $5$ frames,
but in UCT represents $1$ frame. As a result, a UCT rollout of depth
$300$ frames involve the selection of 300 actions, decreasing the
budget of simulated frames by $300$, while the other algorithms need a
branch of $1500$ frames to select $300$ actions, decreasing the budget
by $1500$ frames instead.
}

Table~\ref{performance} shows that both $\IW(1)$ and \BFS\ generally
 outperform \BRFS, which rarely collects any reward in many domains, as 
 the depth of the \BRFS\ search tree results in a lookahead of $0.3$
seconds ($20$ frames or $4$ nodes deep). The notable exception to this
is \Centipede\ where abundant reward can be collected with a shallow
lookahead. On the other hand, both $\IW(1)$ and \BFS\ reach states that 
are up to $350$--$1,500$ frames deep ($70$--$260$ nodes or $6$--$22$ seconds), 
even if $\IW(1)$ does not always use all the simulation frames allocated due to its
pruning. This behaviour can be observed in games such as
\Breakout, \CrazyClimber, \Kangaroo, and \Pooyan, where the average
CPU time for each lookahead is up to $10$ times faster than \BFS. 
Computation time for \UCT\ and \BRFS\ are  similar to \BFS, as the
most expensive part of the computation is the  generation of 
frames through the simulator,  and these three  algorithms 
always use  the full budget.

More interestingly,  \IW(1) outscores \UCT\  in $17$ of the \numgames\ games,
while \BFS\ outscores \UCT\ in $18$.
%\footnote{These numbers can actually
%improve further as 6  games for \IW(1) and 4 games for \BFS\ are still running.}
On the other, \UCT\ does better than $\IW(1)$ and \BFS\ in $9$ and $11$ games respectively.
The relative performance between $\IW(1)$ and \BFS\ is closer with $\IW(1)$ being  the
best of the two in $12$ games, and  \BFS\  in $11$. In terms of the number of games 
where an algorithm is the overall winner, the situation is  even: 
$\IW(1)$ is the overall best in $12$ games, \BFS\ in $11$ games, and \UCT\ in $12$ games. 
Also, \BRFS\ is best in $1$ game (\Centipede), while the other three algorithms
are tied in another game (\Pong).

@@

Likewise, in \Freeway\ and \Berzerk\ both $\IW(1)$ and \BFS\ attain a better
score than the baseline semi-random algorithm \emph{Perturb} in
\cite{bellemare:jair2013}, that beats \UCT\ on those games.
%% to say the 2 lines below must add context: .. eg
%%% ``which in *these games* outperforms both \UCT\ and the reinforcement learning methods.''
%% add it if that's the case ..
\emph{Perturb} is a simple algorithm
that selects  a fixed  action with probability $0.95$, and a random
action with probability $0.05$.  For \emph{Perturb}, BNVB do not report 
the average score  but the  best score.  \emph{Perturb} manages to do well in domains where rewards
are deep but can be reached by  repeating the same action.
This is the case of \Freeway, where a chicken has to run across a ten lane
highway filled with traffic. Every time the chicken gets across, there is one
unit of reward. If the chicken is hit by a car, it has to start again. 
In \Freeway, only $12$ out of $18$ possible actions have an effect:
$6$ actions in moving the chicken up one lane (up-right, up-left, up-fire, up-right-fire, up-left-fire), 
$6$ in moving the chicken down one lane (down-right, down-left, down-fire, down-right-fire, down-left-fire), 
and $6$ actions do nothing.  \emph{Perturb}  does well in this domain 
when the selected fixed action is one of the $6$ that moves the chicken up.  As noted 
in Table~\ref{performance} and seen in the provided video, \UCT\ does not manage to take the chicken
across the highway at all. The reason for \UCT\ to perform erratically 
is that the roll-outs do not manage to collect any reward. A sequence of actions 
with at least $10$ actions embedded that move the chicken up are required in order 
to collect a reward. Since the probability for the roll-out to do an action that moves the chicken up 
is $1/3$, the likelihood of \UCT\ generating such a sequence of actions with the parameters chosen
by BNVB is rather low.
%% the previous number was not correct; no-op actions do not count, and thus needs
%%% 200 more move up actions than move down actions, although this is not accurate either
%%%% as move down actions that take chicken below the screen count as no-ops too ..
$\IW(1)$ does not have this limitation and is best in \Freeway.

@@

$\IW(1)$ also outperforms the best learning algorithm in
\cite{deep-mind-atari} in the $7$ games considered there, and
\BFS\ does so in $6$ of the $7$ games. Comparing 
with the scores reported for the best Reinforcement Learning algorithm
in \cite{bellemare:jair2013}, we note that either $\IW(1)$ and \BFS\ 
are outperforming significantly the best learning algorithm, in those games
where it is reported to perform better than \UCT\ (\Montezuma, \Venture\
and \Bowling). In \Montezuma\ rewards are very
sparse, deep, and most of the actions lead to losing a life with
no immediate penalty or consequence.  All algorithms achieve $0$ score, 
except for \BFS\ that achieves an average score of $80$, and a score
of $400$ in one of the runs. This means that even \BFS\  is not able to 
consistently find rewards in this game. This game and several other games 
like \Breakout, and \SpaceInvaders\ could be much simpler by adding
negative rewards for losing a life. We have indeed observed that
our planning algorithms do not care much about losing lives
until there is just one life left, when their play noticeably improves.
This can be seen in the videos mentioned above, and suggests that
 some simple form of learning would be useful both
for planners and reinforcement learning approaches.

@@

We have shown experimentally that width-based algorithms like $\IW(1)$ and \BFS\
that originate in work in classical planning,  can be used to play  the Atari video
games where they  achieve  state-of-the-art performance. This level of play 
is the result of a structured exploration of the state space that manages to combine 
the scope of blind search algorithms with the efficiency of heuristic search methods. 
There are several  issues worth studying, we  mention two of them.
Regarding planning, the results show the potential of width-based algorithms
for a broader class of planning problems. As shown, indeed, these algorithms
do not require PDDL encodings and can deal with state successor functions
encoded \emph{procedurally} (as in simulators) as long as the structure of the states 
is known. This suggests to us that the algorithms discussed in the 
present paper could be applied in a broad variety of settings where only on-line 
action selection is feasible or suitable.

Regarding the Atari games, we identify two problems in all the planning methods
discussed. First, by using the RAM state of the Atari console one can consider
that they are ``cheating'', as they use information that is not readily available to the human player, 
and second, they do not get better with experience. In the future, we would like to explore 
whether variations of these algorithms  can be used to play from the information conveyed by the  
 state of screen pixels as opposed to the RAM state, as well as exploring the potential uses of the novel
ideas presented for learning, in order to improve performance.
